---
title: "Data Simulations"
author: Tamas Schauer
date: "`r Sys.Date()`"
output: 
        html_document:
                code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

---

##### Asymptotic exponential function
```{r class.source = 'fold-show'}
formulaExpo <- function(x, a, b, c){
        
                y <- c+a*(1-exp(-b*x))
                
                y
}
```

##### Logistic function

```{r class.source = 'fold-show'}
formulaLogi <- function(x, a, b, xmid){
        
                y <- a/(1+exp(-b*(x-xmid)))
                
                y
}
```

##### Generate negative binomial counts

```{r class.source = 'fold-show'}
simCounts <- function(y){
        
        # original in DESeq2: 4/y + 0.1 (too noisy for us)
        dispersion <- 1/y + 0.05
        
        rnbinom(n = length(y), mu = y, size = 1/dispersion)
}

```


## Exponential Asymptote

---

#### Generate data
```{r}
x <- seq(0,45,0.1)

a0 <- 100
b0 <- 0.075
c0 <- 2

y_true <- formulaExpo(x = x, a = a0, b = b0, c = c0)

set.seed(3)
cnts <- simCounts(y_true)
```

#### Fit asymptotic exponential 
```{r}
df <- data.frame(x = x, y = cnts)

fit1 <- nls(y ~ formulaExpo(x, a, b, c), 
            data = df, start = list(a = a0, b = b0, c = c0), 
            control = nls.control(maxiter = 1000))

a1 <- coef(fit1)[1]
b1 <- coef(fit1)[2]
c1 <- coef(fit1)[3]

y_fit1 <- formulaExpo(x = x, a = a1, b = b1, c = c1)
```

#### Fit logistic function

```{r}
fit2 <- nls(y ~ formulaLogi(x, a, b, xmid), 
            data = df, start = list(a = a0, b = b0, xmid = 15), 
            control = nls.control(maxiter = 1000))

a2 <- coef(fit2)[1]
b2 <- coef(fit2)[2]
xmid <- coef(fit2)[3]

y_fit2 <- formulaLogi(x = x, a = a2, b = b2, xmid = xmid)

```

#### Plot

```{r fig.align='center', fig.height=5, fig.width=9}
par(mfrow=c(1,2), mar=c(4,4,1,1), mgp = c(2.25,1,0), oma = c(2,2,2,2))

plot(x, cnts, ylim = c(0, a0*1.5),
     col = "grey", pch = 19, cex = 0.75)

lines(x, y_true, col = "black", lwd = 4)
lines(x, y_fit1, lty=1, col = "blue1", lwd = 2)
lines(x, y_fit2, col = "red3", lwd = 2)

plot.new()
legend("left", lwd = 3,
       legend = c("true","expo fit","logi fit"),
       col = c("black", "blue1","red3"))
```


## Logistic curve

---

#### Generate data

```{r}
x <- seq(0,45,0.1)

a0 <- 100
b0 <- 0.2
xmid = 15

y_true <- formulaLogi(x = x, a = a0, b = b0, xmid = xmid)

set.seed(3)
cnts <- simCounts(y_true)
```

#### Fit asymptotic exponential 

```{r}
df <- data.frame(x = x, y = cnts)

fit1 <- nls(y ~ formulaExpo(x, a, b, c), 
            data = df, start = list(a = a0, b = b0, c = c0), 
            control = nls.control(maxiter = 1000))

a1 <- coef(fit1)[1]
b1 <- coef(fit1)[2]
c1 <- coef(fit1)[3]

y_fit1 <-formulaExpo(x = x, a = a1, b = b1, c = c1)
```

#### Fit logistic function

```{r}
fit2 <- nls(y ~ formulaLogi(x, a, b, xmid), 
            data = df, start = list(a = a0, b = b0, xmid = xmid), 
            control = nls.control(maxiter = 1000))

a2 <- coef(fit2)[1]
b2 <- coef(fit2)[2]
xmid <- coef(fit2)[3]

y_fit2 <- formulaLogi(x = x, a = a2, b = b2, xmid = xmid)

```

#### Plot

```{r fig.align='center', fig.height=5, fig.width=9}
par(mfrow=c(1,2), mar=c(4,4,1,1), mgp = c(2.25,1,0), oma = c(2,2,2,2))

plot(x, cnts, ylim = c(0, a0*1.5),
     col = "grey", pch = 19, cex = 0.75)

lines(x, y_true, col = "black", lwd = 4)
lines(x, y_fit1, lty=1, col = "blue1", lwd = 2)
lines(x, y_fit2, col = "red3", lwd = 2)

plot.new()
legend("left", lwd = 3,
       legend = c("true","expo fit","logi fit"),
       col = c("black", "blue1","red3"))
```

## Inference 1

---

#### Data params (no difference)

```{r}
x <- rep(c(0, 1.5, 3, 6, 12, 25, 45), 3)

a0 <- 2000
b0 <- 0.15
xmid0 = 15

y_true <- formulaLogi(x = x, a = a0, b = b0, xmid = xmid0)

set.seed(5)

cnts1 <- simCounts(y_true)
cnts2 <- simCounts(y_true)

df <- data.frame(x = c(x,x),
                 y = c(cnts1,cnts2),
                 v = rep(c(1,2), each = length(x)))
```

#### Fit full modell

```{r}
fitfull <- nls(y ~ formulaLogi(x, a[v], b[v], xmid[v]),
               data = df,
               start = as.data.frame(rbind(c(a = a0, b = b0, xmid = xmid0),
                                           c(a = a0, b = b0, xmid = xmid0))),
               control =  nls.control(maxiter = 1000, tol = 0.1),
               subset = v %in% 1:2)

summary(fitfull)
```


#### Fit null modell

```{r}
fitnull <- nls(y ~ formulaLogi(x, a, b, xmid),
               data = df,
               start = list(a = a0, b = b0, xmid = xmid0),
               control =  nls.control(maxiter = 1000, tol = 0.1),
               subset = v %in% 1:2)

summary(fitnull)
```


#### NLS without Weights

```{r}
set.seed(4)

pvals <- replicate(n = 1000, expr = {
        
        
        cnts1 <- simCounts(y_true)
        cnts2 <- simCounts(y_true)
        
        df <- data.frame(x = c(x,x),
                         y = c(cnts1,cnts2),
                         v = rep(c(1,2), each = length(x)))
        
        fitv1 <- nls(y ~ formulaLogi(x, a, b, xmid),
                     data =  df,
                     start = list(a = a0, b = b0, xmid = xmid0), 
                     control =  nls.control(maxiter = 1000, tol = 0.1),
                     subset = v == 1)
        
        fitv2 <- nls(y ~ formulaLogi(x, a, b, xmid),
                     data =  df,
                     start = list(a = a0, b = b0, xmid = xmid0), 
                     control =  nls.control(maxiter = 1000, tol = 0.1),
                     subset = v == 2)
        
        fitfull <- nls(y ~ formulaLogi(x, a[v], b[v], xmid[v]), 
                        data = df, 
                        start = as.data.frame(rbind(coef(fitv1), coef(fitv2))), 
                        control =  nls.control(maxiter = 1000, tol = 0.1),
                        subset = v %in% 1:2)
        
        fitnull <- nls(y ~ formulaLogi(x, a, b, xmid),
                        data =  df,
                        start = colMeans(as.data.frame(rbind(coef(fitv1), coef(fitv2)))), 
                        control =  nls.control(maxiter = 1000, tol = 0.1),
                        subset = v %in% 1:2)
        
        anova(fitnull, fitfull)[2,6]
})
```

P-Values should be uniform but they are not

```{r fig.align='center', fig.height=5, fig.width=9}
par(mfrow=c(1,2), mar=c(4,4,1,1), mgp = c(2.25,1,0), oma = c(2,2,2,2))

hist(pvals, breaks = 20)
```


#### Weighted NLS

```{r}
set.seed(5)

pvals <- replicate(n = 1000, expr = {
        
        cnts1 <- simCounts(y_true)
        cnts2 <- simCounts(y_true)
        
        df <- data.frame(x = c(x,x),
                         y = c(cnts1,cnts2),
                         v = rep(c(1,2), each = length(x)))
        
        fitv1 <- nls(y ~ formulaLogi(x, a, b, xmid),
                     data =  df,
                     start = list(a = a0, b = b0, xmid = xmid0), 
                     control =  nls.control(maxiter = 1000, tol = 0.1),
                     subset = v == 1)
        
        
        fitv2 <- nls(y ~ formulaLogi(x, a, b, xmid),
                     data =  df,
                     start = list(a = a0, b = b0, xmid = xmid0), 
                     control =  nls.control(maxiter = 1000, tol = 0.1),
                     subset = v == 2)
        
        fitfull0 <- nls(y ~ formulaLogi(x, a[v], b[v], xmid[v]), 
                        data = df, 
                        start = as.data.frame(rbind(coef(fitv1), coef(fitv2))), 
                        control =  nls.control(maxiter = 1000, tol = 0.1),
                        subset = v %in% 1:2)
        
        df$weights <- 1/(fitted(fitfull0)^2)
        
        fitfull <- nls(y ~ formulaLogi(x, a[v], b[v], xmid[v]), 
                       data = df, 
                       start = as.data.frame(rbind(coef(fitv1), coef(fitv2))), 
                       control =  nls.control(maxiter = 1000, tol = 0.1),
                       weights = weights,
                       subset = v %in% 1:2)
        
        fitnull <- nls(y ~ formulaLogi(x, a, b, xmid),
                       data =  df,
                       start = colMeans(as.data.frame(rbind(coef(fitv1), coef(fitv2)))), 
                       control =  nls.control(maxiter = 1000, tol = 0.1),
                       weights = weights,
                       subset = v %in% 1:2)
        
        anova(fitnull, fitfull)[2,6]
})
```

P-Values should be uniform and they are

```{r fig.align='center', fig.height=5, fig.width=9}
par(mfrow=c(1,2), mar=c(4,4,1,1), mgp = c(2.25,1,0), oma = c(2,2,2,2))

hist(pvals, breaks = 20)
```


## Inference 2

---

#### Data params (with difference in asymptote)

```{r}
x <- rep(c(0, 1.5, 3, 6, 12, 25, 45), 3)

a1 <- 2000
b1 <- 0.18
xmid1 = 15

a2 <- 1000
b2 <- 0.18
xmid2 = 15

y_true1 <- formulaLogi(x = x, a = a1, b = b1, xmid = xmid1)
y_true2 <- formulaLogi(x = x, a = a2, b = b2, xmid = xmid2)

set.seed(9)

cnts1 <- simCounts(y_true1)
cnts2 <- simCounts(y_true2)


df <- data.frame(x = c(x,x),
                 y = c(cnts1,cnts2),
                 v = rep(c(1,2), each = length(x)))

fitv1 <- nls(y ~ formulaLogi(x, a, b, xmid),
             data =  df,
             start = list(a = a1, b = b1, xmid = xmid1), 
             control =  nls.control(maxiter = 1000, tol = 0.1),
             subset = v == 1)


fitv2 <- nls(y ~ formulaLogi(x, a, b, xmid),
             data =  df,
             start = list(a = a2, b = b2, xmid = xmid2), 
             control =  nls.control(maxiter = 1000, tol = 0.1),
             subset = v == 2)
```

```{r fig.align='center', fig.height=5, fig.width=9}
par(mfrow=c(1,2), mar=c(4,4,1,1), mgp = c(2.25,1,0), oma = c(2,2,2,2))


plot(x, cnts1, ylim = c(0,3000),
     col = "grey", pch = 19, cex = 0.75)

lines(seq(0,45,0.01), predict(fitv1, newdata = data.frame(x = seq(0,45,0.01))))

plot(x, cnts2, ylim = c(0,3000),
     col = "grey", pch = 19, cex = 0.75)

lines(seq(0,45,0.01), predict(fitv2, newdata = data.frame(x = seq(0,45,0.01))))
```

#### Test difference of the curves (overall)

```{r}
set.seed(5)

pvals <- replicate(n = 1000, expr = {
        
        cnts1 <- simCounts(y_true1)
        cnts2 <- simCounts(y_true2)
        
        df <- data.frame(x = c(x,x),
                         y = c(cnts1,cnts2),
                         v = rep(c(1,2), each = length(x)))
        
        fitv1 <- nls(y ~ formulaLogi(x, a, b, xmid),
                     data =  df,
                     start = list(a = a1, b = b1, xmid = xmid1), 
                     control =  nls.control(maxiter = 1000, tol = 0.1),
                     subset = v == 1)
        
        fitv2 <- nls(y ~ formulaLogi(x, a, b, xmid),
                     data =  df,
                     start = list(a = a2, b = b2, xmid = xmid2), 
                     control =  nls.control(maxiter = 1000, tol = 0.1),
                     subset = v == 2)
        
        fitfull0 <- nls(y ~ formulaLogi(x, a[v], b[v], xmid[v]), 
                        data = df, 
                        start = as.data.frame(rbind(coef(fitv1), coef(fitv2))), 
                        control =  nls.control(maxiter = 1000, tol = 0.1),
                        subset = v %in% 1:2)
        
        df$weights <- 1/(fitted(fitfull0)^2)
        
        fitfull <- nls(y ~ formulaLogi(x, a[v], b[v], xmid[v]), 
                       data = df, 
                       start = as.data.frame(rbind(coef(fitv1), coef(fitv2))), 
                       control =  nls.control(maxiter = 1000, tol = 0.1),
                       weights = weights,
                       subset = v %in% 1:2)
        
        fitnull <- nls(y ~ formulaLogi(x, a, b, xmid),
                       data =  df,
                       start = colMeans(as.data.frame(rbind(coef(fitv1), coef(fitv2)))), 
                       control =  nls.control(maxiter = 1000, tol = 0.1),
                       weights = weights,
                       subset = v %in% 1:2)
        
        anova(fitnull, fitfull)[2,6]
})
```

P-Values should not be uniform

```{r fig.align='center', fig.height=5, fig.width=9}
par(mfrow=c(1,2), mar=c(4,4,1,1), mgp = c(2.25,1,0), oma = c(2,2,2,2))

hist(pvals, breaks = 20)
```


#### Test only rate parameter (not different)

```{r}
set.seed(4)

pvals <- replicate(n = 1000, expr = {
        
        cnts1 <- simCounts(y_true1)
        cnts2 <- simCounts(y_true2)
        
        df <- data.frame(x = c(x,x),
                         y = c(cnts1,cnts2),
                         v = rep(c(1,2), each = length(x)))
        
        fitv1 <- nls(y ~ formulaLogi(x, a, b, xmid),
                     data =  df,
                     start = list(a = a1, b = b1, xmid = xmid1), 
                     control =  nls.control(maxiter = 1000, tol = 0.1),
                     subset = v == 1)
        
        fitv2 <- nls(y ~ formulaLogi(x, a, b, xmid),
                     data =  df,
                     start = list(a = a2, b = b2, xmid = xmid2), 
                     control =  nls.control(maxiter = 1000, tol = 0.1),
                     subset = v == 2)
        
        fitfull0 <- nls(y ~ formulaLogi(x, a[v], b[v], xmid[v]), 
                        data = df, 
                        start = as.data.frame(rbind(coef(fitv1), coef(fitv2))), 
                        control =  nls.control(maxiter = 1000, tol = 0.1),
                        subset = v %in% 1:2)
        
        df$weights <- 1/(fitted(fitfull0)^2)
        
        fitfull <- nls(y ~ formulaLogi(x, a[v], b[v], xmid[v]), 
                       data = df, 
                       start = as.data.frame(rbind(coef(fitv1), coef(fitv2))), 
                       control =  nls.control(maxiter = 1000, tol = 0.1),
                       weights = weights,
                       subset = v %in% 1:2)
        
        fit2rate <- nls(y ~ formulaLogi(x, a[v], b, xmid[v]), 
                        data = df, 
                        start = c(b = mean(coef(fitv1)[2], coef(fitv2)[2]),
                                  as.data.frame(rbind(coef(fitv1)[-2], coef(fitv2)[-2]))), 
                        weights = weights,
                        control =  nls.control(maxiter = 1000, tol = 0.1),
                        subset = v %in% 1:2)
        
        anova(fit2rate, fitfull)[2,6]
})
```

P-Values should be uniform

```{r fig.align='center', fig.height=5, fig.width=9}
par(mfrow=c(1,2), mar=c(4,4,1,1), mgp = c(2.25,1,0), oma = c(2,2,2,2))

hist(pvals, breaks = 20)
```

#### Test only asymptote parameter (different)

```{r}
set.seed(4)

pvals <- replicate(n = 1000, expr = {
        
        cnts1 <- simCounts(y_true1)
        cnts2 <- simCounts(y_true2)
        
        df <- data.frame(x = c(x,x),
                         y = c(cnts1,cnts2),
                         v = rep(c(1,2), each = length(x)))
        
        fitv1 <- nls(y ~ formulaLogi(x, a, b, xmid),
                     data =  df,
                     start = list(a = a1, b = b1, xmid = xmid1), 
                     control =  nls.control(maxiter = 1000, tol = 0.1),
                     subset = v == 1)
        
        fitv2 <- nls(y ~ formulaLogi(x, a, b, xmid),
                     data =  df,
                     start = list(a = a2, b = b2, xmid = xmid2), 
                     control =  nls.control(maxiter = 1000, tol = 0.1),
                     subset = v == 2)
        
        fitfull0 <- nls(y ~ formulaLogi(x, a[v], b[v], xmid[v]), 
                        data = df, 
                        start = as.data.frame(rbind(coef(fitv1), coef(fitv2))), 
                        control =  nls.control(maxiter = 1000, tol = 0.1),
                        subset = v %in% 1:2)
        
        df$weights <- 1/(fitted(fitfull0)^2)
        
        fitfull <- nls(y ~ formulaLogi(x, a[v], b[v], xmid[v]), 
                       data = df, 
                       start = as.data.frame(rbind(coef(fitv1), coef(fitv2))), 
                       control =  nls.control(maxiter = 1000, tol = 0.1),
                       weights = weights,
                       subset = v %in% 1:2)
        
        fit2asym <- nls(y ~ formulaLogi(x, a, b[v], xmid[v]), 
                        data = df, 
                        start = c(a = mean(coef(fitv1)[1], coef(fitv2)[1]),
                                  as.data.frame(rbind(coef(fitv1)[-1], coef(fitv2)[-1]))), 
                        weights = weights,
                        control =  nls.control(maxiter = 1000, tol = 0.1),
                        subset = v %in% 1:2)
        
        anova(fit2asym, fitfull)[2,6]
})
```

P-Values should not be uniform

```{r fig.align='center', fig.height=5, fig.width=9}
par(mfrow=c(1,2), mar=c(4,4,1,1), mgp = c(2.25,1,0), oma = c(2,2,2,2))

hist(pvals, breaks = 20)

```




